Notes of Topological Transition in a Non-Hermitian Quantum Walk

### 2016-12-07 Notes of various papers/Topological Transition in a Non-Hermitian Quantum Walk/notes.tex
[major]Finish the first version of this notes. There are two arguments that I donot understand.

This paper discusses in general a model charaterised by the follwoing
picture
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{pics/setup.PNG}
    \caption{Setup}
\end{figure}
The equation of motion is
\begin{align}
    i\hbar \dot{\psi}^L_m = \varepsilon_L\psi^L_m + v\psi^R_m
    +v'\psi^R_{m+1} \\
    i\hbar \dot{\psi}^R_m = \tilde{\varepsilon}_R\psi^R_m + v\psi^L_m
    +v'\psi^L_{m-1}
\end{align}

@@

@@

\begin{myquote} \enquote{
our motivation is to provide a simple
model of nuclear spin pumping in spin-blockaded double
quantum dots [9, 10, 11] in the presence of competing
effects of the hyperfine and spin-orbital interactions, as
in Ref.[11].
} \end{myquote}
\section{Anchor}
\label{sec:Anchor}

>>>>

This paper discusses in general a model charaterised by the follwoing
picture
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{pics/setup.PNG}
    \caption{Setup}
\end{figure}
The equation of motion is
\begin{align}
    \begin{aligned}
        \label{eq:eq-of-motion}
    i\hbar \dot{\psi}^L_m = \varepsilon_L\psi^L_m + v\psi^R_m
    +v'\psi^R_{m+1} \\
    i\hbar \dot{\psi}^R_m = \tilde{\varepsilon}_R\psi^R_m + v\psi^L_m
    +v'\psi^L_{m-1}
    \end{aligned}
\end{align}

@@

\textbf{Important Note}: Through out the text we use $m$ as an
abbreviation of $R_m$. So $\psi_m$ means $\psi(R_m)$, $e^{ikm}$ means
$e^{ikR_m}$, and quite decivingly, $e^{ik}$~actually is $e^{ikR_1}$.

@@

\begin{myquote} \enquote{
our motivation is to provide a simple
model of nuclear spin pumping in spin-blockaded double
quantum dots [9, 10, 11] in the presence of competing
effects of the hyperfine and spin-orbital interactions, as
in Ref.[11].
} \end{myquote}

\section{Calculation}
\label{sec:Calculation}

The system starts with the initial state:
\begin{equation}
    \psi^L_m = \delta_{m,0},\quad \psi^R_m = 0
\end{equation}

It evolves according to the following equation of motion
\ref{eq:eq-of-motion}:
\begin{align*}
    \begin{aligned}
    i\hbar \dot{\psi}^L_m = \varepsilon_L\psi^L_m + v\psi^R_m
    +v'\psi^R_{m+1} \\
    i\hbar \dot{\psi}^R_m = \tilde{\varepsilon}_R\psi^R_m + v\psi^L_m
    +v'\psi^L_{m-1}
    \end{aligned}
\end{align*}

The paper seeks to find the following:
\begin{align}
    \braket{\Delta m} &:= \sum_m m P_m \\
    P_m &:= \sum_0^\infty \gamma\abs{\psi^R_m(t)}^2 \dd t
    \label{eq:pm-def}
\end{align}
The result can be obtained both numerically and analytically. First,
the author provides an analytical solution.

\subsection{Analytical Solution}
\label{sec:Analytical-Solution}

\subsubsection{First observation}
\label{sec:First-observation}

The first thing the author points out is that:
\begin{equation}
    \frac{\partial }{\partial t}\abs{\psi}^2
    = \frac{\partial }{\partial t}  \braket{\psi|\psi} \neq 0
\end{equation}
This is caused by one fundamentabl rule of inner product:
\begin{equation}
    \frac{\partial }{\partial t}\braket{f|g} = 
    \braket{\frac{\partial f}{\partial t}|g} +
    \braket{g| \frac{\partial f}{\partial t}}
\end{equation}
The evolution of kets is determined by Schordinger's equation, the
evolution of bras is determined by conjugating the Schordinger's
equation. Thus, it is easy to see:
\begin{equation}
    \label{eq:pt-norm-1}
    i\hbar \frac{\partial }{\partial t}\abs{\psi}^2 =
    i\hbar \frac{\partial }{\partial t}\braket{\psi|\psi} =
    \bra{\psi} H-H^\dagger \ket{\psi}
\end{equation}
Applying the above rules here, we easily find (notice the only
non-hermitian term in the equation of motion \ref{eq:eq-of-motion} is
$\tilde\varepsilon_R = \varepsilon_R -i\hbar \gamma/2$, so
$H-H^\dagger$ is 
$\tilde\varepsilon_R-\tilde\varepsilon_R^*=-i\hbar\gamma$. This also
explains why author choose $\gamma/2$ in $\tilde\varepsilon_R$,
instead of $\gamma$. Thus we have
\begin{align}
    \label{eq:pt-norm-2}
    \frac{\partial }{\partial t}\braket{\psi^R_m|\psi^R_m} = 
    -\gamma \braket{\psi^R_m|\psi^R_m}
\end{align}
For the $L$ sites, the Hamiltonian is hermitian and thus
\begin{align}
    \frac{\partial }{\partial t}\braket{\psi^L_m|\psi^L_m} = 0
\end{align}

Counting them together, we let
\begin{equation}
    \ket{\psi} = \sum_m \ket{\psi^R_m} + \ket{\psi^L_m}
\end{equation}
and find
\begin{equation}
    \frac{\partial }{\partial t}\abs{\psi}^2 =
    \frac{\partial }{\partial t}\braket{\psi|\psi} =
    - \sum_m \gamma\braket{\psi^R_m|\psi^R_m}
\end{equation}

But he the author asserts
\begin{equation}
    \sum_m P_m = 1
\end{equation}
If we perceive that $P_m$ are probabilities, then the above equation
seems logical. But if I write down the formula (by eq \ref{eq:pm-def}):
\begin{equation}
    \sum_m P_m = \sum_m \int_0^\infty \gamma\braket{\psi^R_m|\psi^R_m}
    \overset{?}{=}1
\end{equation}
Then the result is not so obvious.\label{confusion:1}

\subsubsection{Pass to momentum space}
\label{sec:Pass-to-momentum-space}

The basic technique is fourier expansion
\begin{equation}
    \psi = \frac{1}{2\pi} \oint \dd{k} e^{ikm}\psi^R_m
\end{equation}
So the equation of motion \ref{eq:eq-of-motion} become:
\begin{align*}
    i\hbar \frac{\partial }{\partial t}\frac{1}{2\pi}
    \oint \dd{k} e^{ikm} \psi^L_k 
    =\frac{1}{2\pi} \oint \dd{k}  
    \left( \varepsilon_L e^{ikm}\psi^L_k
    + v e^{ikm}\psi^R_k + v' e^{ik(m+1)}\psi^R_k \right)
\end{align*}
similar to the other for $\psi^R_m$.  
The two equations can be simplified to (comparing the terms inside
integration):

\begin{align}
    \label{eq:eq-of-motion-in-k}
    \begin{aligned}
    i\hbar \dot{\psi}^L_k = \varepsilon_L\psi^L_k + v\psi^R_k
    +v' e^{ik} \psi^R_k \\
    i\hbar \dot{\psi}^R_k = \tilde{\varepsilon}_R\psi^R_k + v\psi^L_k
    +v'e^{-ik}\psi^L_{k}
    \end{aligned}
\end{align}

Or in matirx form:
\begin{equation}
    \label{eq:eq-of-motion-in-k-mat}
    i\hbar \frac{\partial }{\partial t}
        \begin{pmatrix}
            \psi^L_k \\ \psi^R_k
        \end{pmatrix}
        =
        \begin{pmatrix}
            \varepsilon_L & A_k \\
            A_k^* & \tilde\varepsilon_R
        \end{pmatrix}
        \begin{pmatrix}
            \psi^L_k \\ \psi^R_k
        \end{pmatrix}
\end{equation}
where $A_k:=v+v'e^{ik}$, and we have asumed $v,v'\in\R$. This is
Eq.(4) in the paper, and can be viewed as an Schordinger's equation in
the momentum space, with $H(k)$ the matrix on the right-hand-side.

By exactly the same technique described from Eq.~\ref{eq:pt-norm-1} to
Eq.~\ref{eq:pt-norm-2}, we can get:
\begin{align*}
    i\hbar\frac{\partial }{\partial t}\braket{\psi^L_k|\psi^L_k} &= 0
    \\
    i\hbar\frac{\partial }{\partial t}\braket{\psi^R_k|\psi^R_k} &= 
        -\gamma\braket{\psi^R_k|\psi^R_k}
\end{align*}

So (let $p_k(t):=\abs{\psi^L_k(t)}^2+\abs{\psi^R_k(t)}^2$):
\begin{equation}
    \frac{\partial }{\partial t}p_k = -\gamma\abs{\psi^R_k(t)}^2
    \label{eq:pt-pk}
\end{equation}

Notice
\begin{equation*}
    -\frac{i}{2\pi}\oint \dd{k} 
    \frac{\partial }{\partial k}(e^{ikm})\ket{\psi^R_k}
    =-i (im) \frac{1}{2\pi}\oint\dd{k} e^{ikm}\ket{\psi^R_k}
    =m\ket{\psi^R_m}
\end{equation*}

Then
\begin{align*}
    m\ket{\psi^R_m} &= -i \frac{1}{2\pi}\oint\dd{k}
    \frac{\partial }{\partial k}(e^{ikm})\ket{\psi^R_k}
    \\
    &=-i\frac{1}{2\pi}\left(
        \eval{e^{ikm}\ket{\psi^R_k}}_{\partial\textrm{BZ}}
        -\oint\dd{k} e^{ikm}\frac{\partial }{\partial k}\ket{\psi^R_k}
    \right)
    \\
    &= i\frac{1}{2\pi}\oint\dd{k}e^{ikm}\frac{\partial }{\partial k}
    \ket{\psi^R_k}
\end{align*}
Hence
\begin{align}
    \braket{\Delta m} &= \sum_m m\int_0^\infty
    \gamma\braket{\psi^R_m|\psi^R_m} 
    = \sum_m \int_0^\infty \gamma
    \bra{\psi^R_m} m\ket{\psi^R_m} \dd t
    \\
    &= \sum_m \int_0^\infty \gamma
    \left(\frac{1}{2\pi}\oint\dd{k} e^{-ikm}\bra{\psi^R_k}\right)
    i\frac{1}{2\pi}\oint\dd{k'}e^{ik'm}\frac{\partial }{\partial k}
        \ket{\psi^R_k} \\
    &= \sum_m \int_0^\infty \gamma
    \frac{i}{(2\pi)^2}\oint\dd{k}\dd{k'} e^{i(k'-k)m}
    \bra{\psi^R_k}\frac{\partial }{\partial k}\ket{\psi^R_k}
\end{align}
Noticing
\begin{equation}
    \sum_m e^{i(k'-k)m} = \delta(k'-k)\frac{(2\pi)^d}{v}
\end{equation}
where $v$ is the volumn of the primitive cell in real lattice and $d$
is the dimention of the real lattice.
So
\begin{align}
    \braket{\Delta m}
    &= \int_0^\infty \frac{i\gamma}{2\pi v}
    \oint\dd{k} \bra{\psi^R_k}
        \frac{\partial }{\partial k}\ket{\psi^R_k}
\end{align}
This same of Eq.(5) in the paper except a constant $v$.
\label{confusion:2}.

Next he sets $\ket{\psi^R_k}=u_k(t)e^{i\theta_k(t)}$, where $u$ is the
length and $\theta$ is the angle, of $\ket{\psi^R_k}$.

The author notes in [18] that the $u_k(t)>0$ for $t>0$, but this is
obvious from Eq.\ref{eq:pt-pk}, which is translated into
\begin{equation}
    \frac{\partial }{\partial t}(u_k)^2 = -\gamma (u_k)^2
\end{equation}
Thus $(u_k)^2$ is an exponential function, which is always positive
unless the initial condition is $0$.

Note
\begin{equation}
    \oint\dd{k} u_k\partial_k u_k 
    = \frac{1}{2}\oint\dd{k}\partial_k (u_k)^2 =
    \frac{1}{2}\eval{(u_k)^2}_{\textrm{\tiny two points of the same value}} = 0
\end{equation}
Then
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{pics/eq7.pdf}
\end{figure}
so
\begin{equation}
    \braket{\Delta m}=\oint\frac{\dd{k}}{2\pi}\int_0^\infty\dd{t}
    (\frac{\partial p_k}{\partial t})
    (\frac{\partial \theta_k(t)}{\partial k})
\end{equation}
This is Eq.7 in the paper, except that I have ignored the $\frac{1}{v}$
difference between my result and the paper's. Then it is straight
forward to obtain
\begin{equation}
    \braket{\Delta m} = \mathcal{I}_0 - \int_0^\infty\dd{t}
    \oint\frac{\dd{k}}{2\pi} p_k \partial_t (\partial_k \theta_k)
    \label{eq:m-I0-int}
\end{equation}
where
\begin{equation}
    \mathcal{I}_0 = \oint\frac{\dd{k}}{2\pi} \left(p_k
    \eval{\frac{\partial \theta_k}{\partial k}}_{t=0}^\infty\right)
\end{equation}
Now the second term in eq.\ref{eq:m-I0-int} can be turned into
(assuming $\theta_k(t)$ is smooth so that order of differentiation
does not matter).
\begin{equation}
    - \int_0^\infty\dd{t}
    \oint\frac{\dd{k}}{2\pi} p_k \partial_t (\partial_k \theta_k)
    =
    \int_0^\infty\dd{t}
    \oint\frac{\dd{k}}{2\pi} 
    \frac{\partial p_k}{\partial k}
    \frac{\partial \theta_k}{\partial t}
    \label{eq:to-be-0}
\end{equation}
This is eq.(9) in the paper.

\subsubsection{Arguing that the Integration \ref{eq:to-be-0} is zero}
\label{sec:Argument-1}

Now the author tries to argue that $p_k$ and $\partial_t\theta_k$ are
even functions of $k$, so that the integration \ref{eq:to-be-0} is $0$.
I cannot follow his argument
\label{confusion:3}. But I find another model, which can found on
\href{https://en.wikipedia.org/wiki/Rabi_cycle}{Wikipedia - Rabi
Cycle}. The Rabi Cycle model can be applied only when $\gamma=0$, i.e.
there is no dissipation. But since the formulae above are continues,
if the integration $p_k$ and $\partial_t\theta_k$ are even, then they
should be even when $\gamma=0$, unless some very strange singular
behavior exist.

\paragraph{$\gamma=0$ case using Two-level system model}$ $

Suppose we have a Hermitian Hamiltonian (let
$\{E_0,W_1,W_2,\Delta\}\subset\R$):
\begin{equation}
    H = E_0\id + W_1\sigma_1+W_2\sigma_2+\Delta\sigma_3 
    = \begin{pmatrix}
        E_0 + \Delta & W_1-iW_2 \\
        W_1+iW_2 & E_0 - \Delta
    \end{pmatrix}
\end{equation}

Supose the system starts at $\ket{\psi(0)}= \begin{pmatrix}
    1\\0
\end{pmatrix}$, then its time evolution will be (Set $\hbar=1$.
Calculation is done in the Wikipedia page, and is checked myself.):
\begin{equation}
    \label{eq:time-evolve-rabi}
    \ket{\psi(t)} = \cos(\theta/2)e^{-iE_+ t}\ket{E_+} - \sin(\theta/2)
    e^{-i\phi}e^{-iE_- t}\ket{E_-}
\end{equation}
where $\ket{E_+}=e^{i\phi/2} \begin{pmatrix}
    \cos(\theta/2) \\ e^{-i\phi}\sin{\theta/2}
\end{pmatrix}$ and $\ket{E_-}= \begin{pmatrix}
    -e^{i\phi}\sin(\theta/2) \\ \cos(\theta/2)
\end{pmatrix}$ are orthonormal eigenkets, and
\begin{align}
\begin{aligned}
    &W:= W_1+iW_2,\quad
    E_{\pm} = E_0\pm \sqrt{\Delta^2+\abs{W}^2}\\
    &\sin(\theta):= \frac{|W|}{\sqrt{\Delta^2+\abs{W}^2}},\,
    \cos(\theta):= \frac{\Delta}{\sqrt{\Delta^2+\abs{W}^2}},\,
\end{aligned}
\end{align}
$\phi$ is such that $W=\abs{W}e^{-i\phi}$.

Here for this paper. Let us suppose for the moment $\gamma=0$, then we
see that $W=A_k^*$, so $W_1 = v+v'\cos{k}$, $W_2=-v'\sin{k}$. 
Also $E_0$ and $\Delta$ are linear combination of
$\varepsilon_R$ and $\varepsilon_L$:
\begin{equation}
    E_0 =\frac{\varepsilon_R+\varepsilon_L}{2},\quad
    \Delta =\frac{\varepsilon_R-\varepsilon_L}{2}
\end{equation}
The system still starts with
$\ket{\psi(0)}$ (since $\psi^L_k(0)=1,\,\psi^R_k(0)=0$). The time
evolved state still follows eq.\ref{eq:time-evolve-rabi}, where the
dependence on $k$ lies in $\sin(\theta),\cos(\theta),$ and $\phi$.
Then by definition, $p_k=\abs{\ket{\psi(t)}}^2$, $u_k(t)$ is the second
component of $\ket{\psi(t)}$.

I was doubtful of the fact that $p_k$ is even in $k$, so I did a
numerical check (see file
"{OneD\_quantum\_walk\_test\_even\_functions\_pk\_uk.m}"), and
unfortunately my guess seems to be wrong. I have given arbitrary
numbers to time $t$, $v$, $v'$, energies $\varepsilon_R$ and
$\varepsilon_L$, and the resulted $p_k$ is always even in $k$, the
resulted $u_k$ is always constantly $0$ in both time variable $t$ and $k$.

This motivates me to calculate it analytically. The results is
contained in files \\
"\texttt{Elementary Calculation of Time Evolution in Two-level model.nb}"

and\\
"\texttt{Elementary Calculation to get pk and uk from Two-Level
model.nb}".

(their pdf version files are also included)

The result is that I can understand $p_k$ is even in $k$, but I cannot
see why $\partial_t u_k$ because its calculation involves getting the
angle of a complex number, which has not good analytical expressions.

\section{Anchor}
\label{sec:Anchor}
